Effectful Semantics for Substructural SSA

### 2023-11-11 pldi24/paper.tex
Tap

\begin{figure}
  \centering
  \begin{subfigure}[t]{0.33\textwidth}
    \begin{BVerbatim}[baseline=t, gobble=3]
    'entry:
      m = 0
      n = 1
      brz i 'exit 'loop
    'loop:
      t = add m n
      m = n
      n = t
      i = sub i 1
      brz i 'exit 'loop
    'exit:
      ret m
    \end{BVerbatim}
    \caption{3-address code}
    \label{fig:3-addr}
  \end{subfigure}
  \begin{subfigure}[t]{0.32\textwidth}
    \begin{BVerbatim}[baseline=t, gobble=3]
      'entry:
      m0 = 0
      n0 = 1
      brz N 
        'exit (m0) 
        'loop (i, m0, n0)
    'loop(i0, m1, n1):
      m2 = n1
      n2 = add m1 n1
      i1 = sub i0 1
      brz z1 
        'exit (m2) 
        'loop (i1, m2, n2)
    'exit(m3):
      ret m3
    \end{BVerbatim}
    \caption{Basic blocks with arguments}
    \label{fig:bb-arg}
  \end{subfigure}
  \begin{subfigure}[t]{0.33\textwidth}
    \begin{BVerbatim}[baseline=t, gobble=3]
    let m0 = 0;
    let n1 = 1;
    if eq i 0 {
      br 'exit (m0)
    } else {
      br 'loop (i, m0, n0)
    } where
    'loop(i0, m1, n1) =>
      let m2 = n1;
      let n2 = add m1 n1;
      let i1 = sub i0 1;
      if eq i1 0 {
        br 'exit (m2)
      } else {
        br 'loop (i1, m2, n2)
      }
    'exit(m3) => ret m3
    \end{BVerbatim}
    \label{fig:our-syn}
    \caption{Our syntax}
  \end{subfigure}
  \caption{A simple program for computing the \texttt{i}th Fibonacci number, represented as 3-address code, basic-blocks with arguments, and in our SSA syntax}
  \Description{}
  \label{fig:ssa-examples}
\end{figure}

\TODO{dominance frontiers?}

\TODO{inductive presentation}

\TODO{seque into what we're doing}

@@

>>>>

\begin{figure}
  \centering
  \begin{subfigure}[t]{0.33\textwidth}
    \begin{BVerbatim}[baseline=t, gobble=3]
    'entry:
      m = 0
      n = 1
      brz i 'exit 'loop
    'loop:
      t = add m n
      m = n
      n = t
      i = sub i 1
      brz i 'exit 'loop
    'exit:
      ret m
    \end{BVerbatim}
    \caption{3-address code}
    \label{fig:3-addr}
  \end{subfigure}
  \begin{subfigure}[t]{0.32\textwidth}
    \begin{BVerbatim}[baseline=t, gobble=3]
      'entry:
      m0 = 0
      n0 = 1
      brz N 
        'exit (m0) 
        'loop (i, m0, n0)
    'loop(i0, m1, n1):
      m2 = n1
      n2 = add m1 n1
      i1 = sub i0 1
      brz z1 
        'exit (m2) 
        'loop (i1, m2, n2)
    'exit(m3):
      ret m3
    \end{BVerbatim}
    \caption{Basic blocks with arguments}
    \label{fig:bb-arg}
  \end{subfigure}
  \begin{subfigure}[t]{0.33\textwidth}
    \begin{BVerbatim}[baseline=t, gobble=3]
    let m0 = 0;
    let n1 = 1;
    if eq i 0 {
      br 'exit (m0)
    } else {
      br 'loop (i, m0, n0)
    } where
    'loop(i0, m1, n1) =>
      let m2 = n1;
      let n2 = add m1 n1;
      let i1 = sub i0 1;
      if eq i1 0 {
        br 'exit (m2)
      } else {
        br 'loop (i1, m2, n2)
      }
    'exit(m3) => ret m3
    \end{BVerbatim}
    \caption{Our syntax}
    \label{fig:our-syn}
  \end{subfigure}
  \caption{A simple program for computing the \texttt{i}th Fibonacci number, represented as 3-address code, basic-blocks with arguments, and in our SSA syntax}
  \Description{}
  \label{fig:ssa-examples}
\end{figure}

For our type theoretic presentation of SSA, we will treat a very slight generalization of the standard basic blocks with arguments syntax, in that we will support \textit{nested scopes} and, in particular, \textit{if-statements}. The program in Figure \ref{fig:bb-arg} translated to our syntax is given in Figure \ref{fig:our-syn}. These do not introduce any new expressivity: an if-statement can be viewed as syntactic sugar for a conditional branch to one of two anonymous basic blocks, while nested scopes simply \textit{restrict} where we can branch to. In particular, in Section \ref{ssec:ssa-norm}, we will formalize interconversion with traditional SSA and show how it is semantically sound to simply merge all nested scopes. The point of the slightly modified syntax is simply to make some of our metatheoretic results more practical to state.

@@

\subsection{Normalization}

\label{ssec:ssa-norm}
